Variable Selection in Convex Function Estimation

### 2014-04-18 tex/scam/sparsistency.tex
paper update

Our analysis then differs from the now-standard ``primal-dual witness
technique''~\cite{Wain:09a}. Primal-dual witness explicitly solves all the dual variables, but because our optimization is more complex, we do not solve the dual variables on $S$; we instead write the dual variables on $S^c$ as a function of the restricted regression \emph{residual}, which is implicitly a function of the dual variables on $S$.

\begin{theorem} (Deterministic setting)
\label{thm:deterministic}
Let $\{\hat{d}_k, \hat{c}_k\}_{k \in S}$ be the minimizer of the restricted regression, that is, the solution to optimization (\ref{opt:alternate_opt}) where we restrict $k \in S$. Let $\hat{d}_k = 0$ and $ \hat{c}_k = 0$ for $k \in S^c$.
Let $\hat{r} \coloneqq Y - \bar{Y}\mathbf{1}_n - \sum_{k \in S} (\Delta_k \hat{d}_k -
\hat{c}_k \mathbf{1}_n)$ be the restricted regression residual. For $k
\in \{1,...,p\}$, Let $\Delta_{k, j} \in R^n$ be the $j$-th column of $\Delta_k$, i.e. $\max( X_k - X_{k (j)} \mathbf{1}_n, 0)$. \\

Suppose for all $j$ and all $k\in S^c$, $\lambda_n > | \frac{1}{n}
\hat{r}^\tran \Delta_{k,j}|$. Then $\hat{\mu}$ and $\hat{d}_k, \hat{c}_k$ for $k=1,...,p$ is an optimal solution to the full regression \ref{opt:alternate_opt}. Furthermore, any solution to the optimization program \ref{opt:alternate_opt} must be zero on $S^c$.
\end{theorem}

This result holds regardless of whether we impose the boundedness and Lipschitz conditions in optimization~\ref{opt:alternate_opt}.
The full proof of Theorem~\ref{thm:deterministic} is in Section~\ref{sec:deterministic_proof} of the Appendix.

\begin{remark}
  The incoherence condition of \cite{Wain:09a} is implicitly encoded
  in our condition on $\lambda_n, \hat{r}, \Delta_{k,j}$. We can
  reconstruct the incoherence condition if we assume that the true
  function $f_0$ is linear and that our fitted functions $\hat{f}_k$
  are linear as well.
\end{remark}

Theorem~\ref{thm:deterministic} allows us to analyze false negative
rates and false positive rates separately. To control false positives,
we study when the condition $\lambda_n > | \frac{1}{n} \hat{r}^\tran
\Delta_{k,j}|$ is fulfilled for all $j$ and all $k \in S^c$. To
control false negatives, we study the restricted regression.

@@

\begin{packed_enum}
\item Let $F$ be a distribution supported and positive on $\mathcal{X}=[-b,b]^p$. Let $X^{(1)},..., X^{(n)} \sim F$ be iid.
\item Let $Y = f_0(X) + \epsilon$ where $\epsilon$ is zero-mean noise. Let $Y^{(1)},...,Y^{(n)}$ be iid.
\item Let $S = \{1,...,s\}$ denote the relevant variables where $s\leq p$, i.e.,
  $f_0(X) = f_0(X_S)$.
\item Let $f^*_1,...,f^*_s \coloneqq \argmin_{f_1,...,f_s} \{ \E\Big(f_0(X) -\E f_0(X) - \sum_{k=1}^s f_k(X_k)\Big)^2 \,|\, \E[f_k(X_k)] = 0 \}$.
\end{packed_enum}

Each of our theorems will use a subset of the following assumptions:
\begin{packed_enum}
\item[A1:] $X_S, X_{S^c}$ are independent.  \ A1': $\{ X_k \}_{k \in S}$ are independent.
\item[A2:] $\|f_0\|_\infty \leq sB$ \  A2': $f_0$ is convex,
  twice-differentiable, and $L$-Lipschitz.
\item[A3:] Suppose $\epsilon$ is mean-zero sub-Gaussian, independent of $X$, with sub-Gaussian scale $\sigma$, i.e. for all $t \in \R$, $\E e^{t \epsilon} \leq e^{\sigma^2 t^2 / 2}$.
\item[A4:] For all $k=1,...,s$, $\E(f^*_s(X_k))^2 \geq \alpha$ for some positive constant $\alpha$.
\end{packed_enum}

We will use assumptions A1, A2, A3 to control the probability of false
positives and the stronger assumptions A1', A2', A3, A4 to control the
probability of false negatives.  Assumption A4 can be
weakened so that the relevant functions satisfy
$\E(f^*_s(X_k))^2 \geq \alpha_n$ for $\alpha_n$ decaying to zero 
at an appropriate rate.

\begin{remark}
  We make strong assumptions on the covariates in A1 in order to make
  very weak assumptions on the true regression function $f_0$ in
  A2. In particular, we do not assume that $f_0$ is additive. Relaxing
  these assumptions is an interesting direction for future work. 
  %Strong assumptions on the covariates are not uncommon in nonparametric
  %variable selection analysis~\cite{lafferty2008rodeo}. 
  %[TODO: refer to correlated design experiment].
\end{remark}

\begin{remark}
Assumption A4 ensures that the relevant variables are ``relevant enough''. Under A4, the population risk of an additive function with $s-1$ components is at least $\alpha$ larger than the population risk of the optimal additive function with $s$ components. See lemma~\ref{lem:minus_one_risk_increase} in section~\ref{sec:false_negative_proof} of the appendix.
\end{remark}

\begin{theorem} (Controlling false positives) 
\label{thm:false_positive}
Suppose assumptions A1, A2, A3 hold. Suppose also that we run optimization~\eqref{opt:alternate_opt} with the $B$-boundness constraint. Let $c,C$ be absolute constants.
Suppose $\lambda_n \geq c b (sB + \sigma) \sqrt{ \frac{s}{n} \log n
  \log (pn)}$.  Then with probability at least $ 1 - \frac{C}{n}$, for all $j,k$, $\lambda_n >  | \frac{1}{n} \hat{r}^\tran \Delta_{k,j}|$.
Therefore, any solution to the full regression (\ref{opt:alternate_opt}), with boundedness constraint, is zero on $S^c$. 
\end{theorem}

The proof of Theorem~\ref{thm:false_positive} exploits independence of
$\hat{r}$ and $\Delta_{k,j}$ from A1, and then uses concentration of
measure results to argue that $| \frac{1}{n} \hat{r}^\tran
\Delta_{k,j}|$ concentrates around zero at a desired rate. The fact
that $\hat{r}$ is a centered vector is crucial to our proof, and our
theory thus further illustrates the importance of imposing the
centering constraints in optimization \eqref{opt:alternate_opt}. Our
proof uses the concentration of the average of
data sampled \emph{without} replacement
\cite{serfling1974probability}. The full proof of
Theorem~\ref{thm:false_positive} is in
Section~\ref{sec:false_positive_proof} of the Appendix.

\begin{theorem} (Controlling false negatives)
\label{thm:false_negative}
Suppose assumptions A1', A2', A3, A4 hold. Let $\ds \hat{f} = \{ \hat{d}_k, \hat{c}_k\}_{k\in S}$ be any solution to the restricted regression with both the $B$-boundedness and $L$-Lipschitz constraint. Let $c,C$ be absolute constants.
Suppose $s L \lambda_n \rightarrow 0$ and $Lb(sB+\sigma)sB \sqrt{\frac{s}{n^{4/5}} \log sn} \rightarrow 0$.
Then, for sufficiently large $n$, $\hat{f}_k = (\hat{d}_k, \hat{c}_k)
\neq 0$ for all $k \in S$ with probability at least $1-\frac{C}{n}$.
\end{theorem}

@@

Combining Theorem~\ref{thm:false_positive} and
~\ref{thm:false_negative} and ignoring dependencies on $b,B,L,\sigma$,
we have the following result.
\begin{corollary}
  Assume A1', A2', A3, A4. Let $\lambda_n = \Theta\left( \sqrt{
  \frac{s^3}{n} \log n \log (pn)} \right)$. Suppose $s \lambda_n
  \rightarrow 0$ and $\sqrt{\frac{s^5}{n^{4/5}} \log sn} \rightarrow
  0$. Let $\hat{f_n}$ be a solution to (\ref{opt:alternate_opt}) with
  boundedness and Lipschitz constraints. Then 
  $\P( \trm{supp}(\hat{f_n}) = \trm{supp}(f_0) ) \rightarrow 1$.
\end{corollary}
The above corollary implies that sparsistency is achievable at the same exponential scaling of the ambient dimension $p = O(\exp(n^c)), c<1$ rate as parametric models. The cost of nonparametric modeling is reflected in the scaling with respect to $s$, which can only scale at $o(n^{4/25})$.

>>>>

Our analysis then differs from the now-standard ``primal-dual witness
technique''~\cite{Wain:09a}. Primal-dual witness explicitly solves all the dual variables, but because our optimization program is more complex, we do not solve the dual variables on $S$; we instead write the dual variables on $S^c$ as a function of the restricted regression \emph{residual}, which is implicitly a function of the dual variables on $S$.

\begin{theorem} (Deterministic setting)
\label{thm:deterministic}
Let $\{\hat{d}_k \}_{k \in S}$ be a minimizer of the restricted regression, that is, the solution to optimization (\ref{opt:alternate_opt}) where we restrict $k \in S$. 
Let $\hat{r} \coloneqq Y - \sum_{k \in S} \bar{\Delta}_k \hat{d}_k$ be the restricted regression residual. \\

Suppose for all $k\in S^c$, for all $i=1,...,n$, $\lambda_n > | \frac{1}{2n}
\hat{r}^\tran \mathbf{1}_{(i:n)}|$ where $\mathbf{1}_{(i:n)}$ is 1 on the coordinates of the $i$-th largest to the $n$-th largest entries of $X_k$ and 0 elsewhere.\\

Then the following are true:
\begin{enumerate}
\item Let $\hat{d}_k = 0$ for $k \in S^c$, then \{$\hat{d}_k\}_{k=1,...,p}$ is an optimal solution to optimization~\ref{opt:alternate_opt}. Furthermore, any solution to the optimization program \ref{opt:alternate_opt} must be zero on $S^c$.
\item For all $k \in S^c$, the solution to optimization~\ref{opt:alternate_opt_concave} must be 0 and must be unique.
\end{enumerate}

\end{theorem}

This result holds regardless of whether we impose the Lipschitz conditions in optimization~\ref{opt:alternate_opt} and~\ref{opt:alternate_opt_concave} or not.
The full proof of Theorem~\ref{thm:deterministic} is in Section~\ref{sec:deterministic_proof} of the Appendix.\\

Theorem~\ref{thm:deterministic} allows us to analyze false negative
rates and false positive rates separately. To control false positives,
we analyze when the condition on $\lambda_n$ is fulfilled for all $j$ and all $k \in S^c$. To control false negatives, we analyze the restricted regression.

@@

\begin{packed_enum}
\item Let $F$ be a distribution supported and positive on $\mathcal{X}=[-b,b]^p$. Let $X^{(1)},..., X^{(n)} \sim F$ be iid.
\item Let $Y = f_0(X) + w$ where $w$ is zero-mean noise. Let $Y^{(1)},...,Y^{(n)}$ be iid.
\item Let $S_0 = \{1,...,s_0\}$ denote the relevant variables where $s_0 \leq p$, i.e.,
  $f_0(X) = f_0(X_{S_0})$.
\end{packed_enum}

Let $\mathcal{C}^1$ denote the set of univariate convex functions supported on $[-b,b]$. Define $\mathcal{C}^{p}$ as the set of convex additive function
$\mathcal{C}^p \equiv \{ f \,:\, f = \sum_{k=1}^p f_k, \,
   f_k \in \mathcal{C}^1 \} $
Let $f^*(x) = \sum_{k=1}^p f^*_k(x_k)$ be the population risk minimizer:
$f^* \equiv \arg\min_{f \in \mathcal{C}^p} \E\big(f_0(X) - f^*(X) \big)^2$.\\

Similarly, we define $\mh \mathcal{C}^1$ as the set of univariate concave functions supported on $[-b, b]$ and let 
$g^*_k = \arg\min_{g_k \in \mh \mathcal{C}^1} \E \big( f_0(X) - f^*(X) - g_k(X_k) \big)^2$. \\

We let $S = \{ k = 1,...,p \,:\, f^*_k \neq 0 \trm{ or } g^*_k \neq 0\}$. By additive faithfulness (theorem~\ref{thm:acdc_faithful}), it must be that $S_0 \subset S$.

Each of our theorems will use a subset of the following assumptions:
\begin{packed_enum}
\item[A1:] $X_S, X_{S^c}$ are independent. 
\item[A2:] $f_0$ is convex and twice-differentiable. 
\item[A3:] $|\partial_{x_k} f_0 | \leq L$ for all $k$
\item[A4:] $w$ is mean-zero subgaussian, independent of $X$, with subgaussian scale $\sigma$, i.e. for all $t \in \R$, $\E e^{t \epsilon} \leq e^{\sigma^2 t^2 / 2}$.
\end{packed_enum}
By assumption A1, $f^*_k$ is must be zero for $k\notin \{1,...,s\}$.

We define $\alpha_f, \alpha_g$ as a measure of the signal strength of the weakest variable:
\begin{align*}
\alpha_f &= \inf_{f \in \mathcal{C}^p \,:\, \exists k ,\, f^*_k \neq 0 \,\wedge\, f_k = 0} 
       \Big\{ \mathbb{E} \big( f_0(X) - f(X) \big)^2 - 
        \mathbb{E} \big( f_0(X) - f^*(X) \big)^2  \Big\}\\
\alpha_g &=   \min_{k \in S \,:\, g^*_k \neq 0}
      \Big\{ \mathbb{E} \big( f_0(X) - f^*(X) \big)^2 - 
    \mathbb{E} \big( f_0(X) - f^*(X) - g^*_k(X_k) \big)^2 \Big\}
\end{align*}

Intuitively, if $\alpha_f$ is smaller, then it is easier to make a false omission in the additive convex stage of the procedure. If $\alpha_g$ is smaller, then it is easier to make a false omission in the decoupled concave stage of the procedure.

\begin{remark}
  We make strong assumptions on the covariates in A1 in order to make
  very weak assumptions on the true regression function $f_0$ in
  A2. In particular, we do not assume that $f_0$ is additive. 
  %Strong assumptions on the covariates are not uncommon in nonparametric
  %variable selection analysis~\cite{lafferty2008rodeo}. 
  %[TODO: refer to correlated design experiment].
\end{remark}

\begin{theorem} (Controlling false positives) \\
\label{thm:false_positive}
Suppose assumptions A1-A4 hold.

Suppose $\lambda_n \geq \mathbf{c} s Lb \sigma  \sqrt{ \frac{1}{n} \log^2 np}$, then with probability at least $ 1 - \frac{C}{n}$, for all $k \in S^c$, and for all $i'=1,...,n$:
\[
\lambda_n > \Big| \frac{1}{2n}\hat{r}^\tran \mathbf{1}_{(i':n)} \Big|
\]

And therefore，for all $k \in S^c$, both the AC solution $\hat{f}_k$, from optimization~\ref{opt:alternate_opt}, and the DC solution $\hat{g}_k$, from optimization~\ref{opt:alternate_opt_concave} are zero. \\
\end{theorem}

The proof of Theorem~\ref{thm:false_positive} exploits independence of
$\hat{r}$ and $X_k$ from A1; when $\hat{r}$ and $X_k$ are independent, $\hat{r}^\tran \mathbf{1}_{(i':n)}$ is the sum of $n - i' +1$ random coordinates of $\hat{r}$.  We can then uses the concentration of measure result for sampling without replacement to argue that $| \frac{1}{n} \hat{r}^\tran\mathbf{1}_{(i':n)}|$ is small with high probability. The full proof of
Theorem~\ref{thm:false_positive} is in
Section~\ref{sec:false_positive_proof} of the Appendix.

\begin{theorem} (Controlling false negatives) \\
\label{thm:false_negative}
Suppose assumptions A1-A4 hold. Let $\hat{f}$ be any AC solution to the restricted regression with $L$-Lipschitz constraint and let $\hat{g}_k$'s be any DC solution to the restricted regression with $L$-Lipschitz constraint.\\

Suppose $\lambda \leq \mathbf{c} sLb  \sqrt{\frac{1}{n} \log^2 np}$ and suppose $n$ is large enough such that $(Lb)^3 \sigma \sqrt{\frac{s^5}{n^{4/5}} \log^2 np} \leq 1$.\\

Suppose 
$\frac{\alpha_f}{\sigma} \geq \mathbf{c} (Lb)^3 \sqrt{\frac{s^5}{n^{4/5}} \log^2 np}$ and $\frac{\alpha_g^2}{\sigma} \geq \mathbf{c} (Lb)^3 \sqrt{\frac{s^5}{n^{4/5}} \log^2 2np}$.\\

Then, with probability at least $1 - \frac{1}{n}$, we have that for all $k \in S$, $\hat{f}_k \neq 0$ or $\hat{g}_k \neq 0$.

\end{theorem}

@@

Combining Theorem~\ref{thm:false_positive} and
~\ref{thm:false_negative} 
we have the following result.
\begin{corollary}
Suppose assumptions A1-A4 hold. Let $\hat{f}$ be any AC solution and $\hat{g}_k$'s be any DC solution, both with $L$-Lipschitz constraints, both with 
$\lambda = \Theta \Big( sLb \sqrt{\frac{1}{n} \log^2 np} \Big)$. \\

Suppose 
$\frac{\alpha_f}{\sigma} \geq \mathbf{c} (Lb)^3 \sqrt{\frac{s^5}{n^{4/5}} \log^2 np}$ and $\frac{\alpha_g^2}{\sigma} \geq \mathbf{c} (Lb)^3 \sqrt{\frac{s^5}{n^{4/5}} \log^2 2np}$.\\

Then, for all large enough $n$, we have that with probability at least $1-\frac{1}{n}$:
\begin{align*}
\hat{f}_k \neq 0 \trm{ or } \hat{g}_k \neq 0 &\trm{ for all } k \in S\\
\hat{f}_k = 0 \trm{ and } \hat{g}_k = 0 & \trm{ for all } k \notin S
\end{align*}

\end{corollary}
The above corollary implies that sparsistency is achievable at the same exponential scaling of the ambient dimension $p = O(\exp(n^c)), c<1$ rate as parametric models. The cost of nonparametric modeling is reflected in the scaling with respect to $s$, which can only scale at $o(n^{4/25})$.
